\newcommand{\teta}{\tilde{\eta}}

Incompressible flow represents the zero-Mach number limit of fluid
flow---no compressibility effects are modeled.  We can extend the
ideas of incompressibe flow to allow us to model some compressibility
effects, giving rise to low Mach number methods.

\section{Low Mach divergence constraints}

The key idea in solvers for low Mach number flows is that, as a fluid
element advects, its pressure remains the same as the background
state.  For an atmosphere, this background state is a hydrostatic
profile.  For a smallscale combustion flow, this background state is a
spatially constant pressure (and constaint-in-time if the domain is
open).

We can derive the constraint on the velocity field by considering the
Lagrangian derivative of the pressure---this captures the change in
pressure of a fluid element as it moves through the domain.



\section{Extending the incompressible algorithm}

\subsection{Variable-coefficient elliptic equation}

We now need to solve an elliptic equation of the form:
\begin{equation}
\nabla \cdot (\eta \nabla \phi) = f
\end{equation}

If we denote the discrete divergence and gradient operators as $D$ and $G$,
then our operator will be $L_\eta \equiv D \eta G$.  If we wish to
use a cell-centered discretization for $\phi$, then using a standard 
centered-difference for $D$ and $G$ will result in a stencil that reaches
two zones on either side of the current zone.  This can lead to an
odd-even decoupling. \MarginPar{cite the appropriate Bell paper}

\MarginPar{there is a Pao and Colella (or Pao's thesis?) that also 
discusses issues with cell-centered}

Instead, we again use an approximate projection.  We discretize the
variable-coefficient Laplacian as:
\begin{align}
(L_\eta \phi)_{i,j} = 
 & \frac{\eta_{i+1/2,j} (\phi_{i+1,j} - \phi_{i,j}) -
        \eta_{i-1/2,j} (\phi_{i,j} - \phi_{i-1,j})}{\Delta x^2} + \\
 & \frac{\eta_{i,j+1/2} (\phi_{i,j+1} - \phi_{i,j}) -
        \eta_{i,j-1/2} (\phi_{i,j} - \phi_{i,j-1})}{\Delta y^2}
\end{align}

We can define the interface values of $\eta$ as the averages of the
cell-centered values.  Our elliptic equation is then 
\begin{equation}
(L_\eta \phi)_{i,j} = f_{i,j}
\end{equation}

The relaxation method for this operator again relies on isolating
$\phi_{i,j}$, yielding:
\begin{equation}
\phi_{i,j} = \frac{\teta_{i+1/2,j}\phi_{i+1,j} + \teta_{i-1/2,j}\phi_{i-1,j} +
                   \teta_{i,j+1/2}\phi_{i,j+1} + \teta_{i,j-1/2}\phi_{i,j-1} -
                   f_{i,j} } 
                  {\teta_{i+1/2,j} + \teta_{i-1/2,j} + 
                   \teta_{i,j+1/2} + \teta_{i,j-1/2}}
\end{equation}
with the shorthand that $\teta_{i\pm1/2,j} = \teta_{i\pm1/2,j}/\Delta x^2$
and $\teta_{i,j\pm1/2} = \teta_{i,j\pm1/2}/\Delta y^2$.

\section{Reactive flows}

Our constraint equation is $\nabla \cdot U = S$.  Decomposing the velocity field
as 
\begin{equation}
U^\star = U^d + \frac{1}{\rho} \nabla \phi
\end{equation}
our Poisson equation can be defined by taking the divergence, and
using $\nabla \cdot U^d = S$, giving
\begin{equation}
\nabla \cdot \frac{1}{\rho} \nabla \phi = \nabla \cdot U^\star - S
\end{equation}


\section{Atmospheric flows}

Our constraint equation is $\nabla \cdot (\beta_0 U) = 0$.  Decomposing the velocity field
as 
\begin{equation}
U^\star = U^d + \frac{1}{\rho} \nabla \phi
\end{equation}
our Poisson equation can be defined by multiplying by $\beta_0$ and
taking the divergence, and using $\nabla \cdot (\beta_0 U^d) = 0$, giving
\begin{equation}
\nabla \cdot \frac{\beta_0}{\rho} \nabla \phi = \nabla \cdot (\beta_0 U^\star)
\end{equation}
